Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa

### 2022-05-12 src/docs_12-05-22-mrc-gida/12-05-22-mrc-gida.Rmd
Final updates to MRC presentation, including a plot highlighting girls cohabiting in MOZ

@@

* Prevention packages are made up of a core package, available to everyone, and more intensified interventions
* To have the greatest impact, it's important to priorisite intensified interventions to those at highest risk
* The Global AIDS Strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon
  1. Population-level HIV incidence in the local population
  2. Individual-level sexual risk behaviour

@@

1. Enable implementation of prevention strategies stratified by incidence and behaviour
2. Assess the benefits of such approaches

@@

* We used sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys to place respondents into four risk groups:
  1. $k = 1$: Not sexually active
  2. $k = 2$: One cohabiting sexual partner
  3. $k = 3$: Non-regular sexual partner(s)
  4. $k = 4$: Female sex workers
* District-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]
* Risk ratios from ALPHA network analysis [@slaymaker2020croi] and UNAIDS analysis led by Keith Sabin

@@

* Multinomial logistic regression model for the proportion of AGYW in the $k = 1, 2, 3^{+}$ risk groups, using all 47 surveys
* Selected model included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Correlated temporal effects (AR1) 
* Multinomial-Poisson transformation allowed us to use `R-INLA` for inference

@@

* Logistic regression model for the proportion of those in the $k = 3^{+} = \{3, 4\}$ risk groups who are in the $k = 4$ risk group, using only the 13 surveys with a specific transactional sex question
* Selected model included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Clients of FSW covariates [@hodgins2022population]
* Also used `R-INLA` for inference here

@@

* Integration of all relevant surveys
  * Modular approach allowed estimating FSW proportion even for surveys without a specific transactional sex question
* Alleivating small-sample sizes (that caused direct estimates to display large variance)
  * We borrowed information across space, between countries and over surveys so that our estimates more plausibly reflect spatial heterogeneity
* Estimates where there isn't direct data
  * Although some people think of this as "making up data", the data almost never "speak for themselves" (everything is a model)
  * Uncertainty should be higher in regions with infilling: important to transparently communicate this

@@

* We used our risk group proportion estimates and relative risk ratio estimates from the literature to disaggregate Naomi general population HIV incidence estimates
\begin{align*}
I_{ia} &= \sum_k \lambda_{iak} N_{iak} \\
&= \sum_k \lambda_{ia2} \text{RR}_k N_{iak}.
\end{align*}

@@

```{r echo=FALSE, out.width = "400px"}
knitr::include_graphics("depends/infections-reached.png")
```

@@

* Simplistic infections reached analysis
  * No accounting for difficulty in reaching each strata
  * Variable intervention effectiveness
  * Change in strata membership
* Under-reporting of high risk sexual behaviours
  * Variation in under-reporting (likely by age, foremost, and location, less so) particularly concerning
* Simple risk groups, not clearly linked to data
  * Didn't consider other important characteristics that may determine risk e.g. condom usage
* Only focused on AGYW 15-29
  * Could be extended to adults of both sexes aged 15-49

## Policy takeaways

* Our risk group estimates can be used to help countries implement the Global AIDS Strategy (user guide currently under development)
* It's important to reach FSW (not a new message!)
* Countries have different epidemic profiles

## Methods takeaways

* `R-INLA` and the multinomial-Poisson trick are a good choice for spatio-temporal multinomial models
  * The `group` and `replicate` features can be used to define quite complex Kronecker product effects
  * However, can run into limitations on model complexity and I might be moving to something more flexible e.g. `tmb`, `aghq` to continue development of this project

@@

* This project was joint work with members of the HIV inference group, particularly Katie Risher and Jeff Eaton
  * \faHome $\,$ `hiv-inference.org`
* You can find me via
  * \faEnvelope $\,$ `ath19@ic.ac.uk`
  * \faHome $\,$ `athowes.github.io`

>>>>

## Prevention packages

* Prevention
  * Core package
  * Intensified interventions
* It's important to priorisite intensified interventions to those at highest risk

@@

::: columns

:::: column

The Global AIDS strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon

1. Population-level HIV incidence
2. Individual-level sexual risk behaviour

::::

:::: column

```{r echo=FALSE, out.width = "125px", fig.cap="Global AIDS strategy"}
knitr::include_graphics("global-aids-strategy.png")
```

::::

:::

@@

1. Enable implementation of prevention stratified by incidence and behaviour
2. Assess the benefits of such approaches

@@

* We used sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys to place respondents into four risk groups:
  1. Not sexually active
  2. One cohabiting sexual partner
  3. Non-regular sexual partner(s)
  4. Female sex workers
* District-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]
* Risk ratios from ALPHA network analysis [@slaymaker2020croi] and UNAIDS analysis led by Keith Sabin

@@

* Multinomial logistic regression model for the proportion of AGYW in the $k = 1, 2, 3^{+}$ risk groups, using all 47 surveys
* Selected model (by CPO) included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Correlated temporal effects (AR1) 
* Multinomial-Poisson transformation allowed use of `R-INLA` for inference

@@

* Logistic regression model for the proportion of those in the $k = 3^{+} = \{3, 4\}$ risk groups who are in the $k = 4$ risk group, using only the 13 surveys with a specific transactional sex question
* Selected model (by CPO) included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Clients of FSW covariates [@hodgins2022population]
* Used `R-INLA` for inference

@@

* Integration of all relevant surveys
  <!-- * Modular approach allowed estimating FSW proportion even for surveys without a specific transactional sex question -->
* Alleivating small-sample sizes by borrowing information
  <!-- * We borrowed information across space, between countries and over surveys so that our estimates more plausibly reflect spatial heterogeneity -->
* Estimates where there isn't direct data
  <!-- * Although some people think of this as "making up data", the data almost never "speak for themselves" (everything is a model) -->
  <!-- * Uncertainty should be higher in regions with infilling: important to transparently communicate this -->

@@

* Risk group proportion estimates plus relative risk ratio estimates to disaggregate general population HIV incidence estimates
\begin{align*}
I_{ia} &= \sum_k \lambda_{iak} N_{iak} \\
&= \sum_k \lambda_{ia2} \text{RR}_k N_{iak}.
\end{align*}

@@

```{r echo=FALSE, out.width = "350px", fig.cap = "New infections reached prioritising according to different stratifications."}
knitr::include_graphics("depends/infections-reached.png")
```

```{r echo=FALSE, out.width = "350px", fig.cap = "As previous slide but specific to Mozambique."}
knitr::include_graphics("depends/infections-reached-moz.png")
```

```{r echo=FALSE, out.width = "300px", fig.cap = "Mozambique stands out."}
knitr::include_graphics("depends/cohabiting-girls.pdf")
```

@@

* Simplistic infections reached analysis
  <!-- * No accounting for difficulty in reaching each strata -->
  <!-- * Variable intervention effectiveness -->
  <!-- * Change in strata membership -->
* Under-reporting of high risk sexual behaviours
  <!-- * Variation in under-reporting (likely by age, foremost, and location, less so) particularly concerning -->
* Risk groups definition justification not clear
  <!-- * Didn't consider other important characteristics that may determine risk e.g. condom usage -->
* Only focused on AGYW 15-29
  <!-- * Could be extended to adults of both sexes aged 15-49 -->

## Takeaways

* Risk group estimates can help implement the Global AIDS Strategy; tool and user guide currently being prepared!
* Importance of reaching FSW
* Countries have different epidemic profiles

@@

* Joint work with members of the HIV inference group (`hiv-inference.org`) particularly Katie Risher and Jeff Eaton
* The code for this project is at `github.com/athowes/multi-agyw`
* You can find me online at `athowes.github.io`
